Flexible Binary Instrumentation for the Kernel

### 2013-11-13 VEE-14/paper.tex
Working on characterizing RCU bugs

Read-Copy-Update (RCU) is a scalable, lock-free synchronization mechanism used in the Linux kernel \cite{RCU,RCUInLinux}. RCU is used to synchronize accesses (reads) and modifications (writes) of shared state. When correctly used, RCU permits concurrent reads and writes to shared data structures, but only if the following properties are satisfied: \begin{enumerate}
	\item Readers must tolerate an inconsistent view of the shared data structure. That is, a reader operating concurrently with a writer might observe either the version of the data before or after the writer's changes have been published.
	\item Readers must operate within \emph{read-side} critical sections, which are delimited by \texttt{rcu\_read\_lock} and \texttt{rcu\_read\_unlock}, respectively.
	\item Within read-side critical sections, readers can only read a shared data structure after obtaining a pointer to that shared data structure using the \texttt{rcu\_dereference} API function. \texttt{rcu\_dereference} operates on a pointer to a shared data structure, and returns a pointer to that can be used to safely access the shared data structure within the remainder of the reader's read-side critical section. Notably, accessing the original data structure and not the dereferenced data structure pointer--even after an \texttt{rcu\_dereference}--is undefined behavior. That is, within a read-side critical section, RCU only guarantees the existence of data structures pointed to by RCU-dereferenced pointers.
	\item Writers must synchronize among themselves, e.g. via mutual exclusion locks or spin locks, before any writes to a shared data structure are performed. We say writes to the shared data structure occur within \emph{write-side} critical sections.
	\item RCU only defines how writers can update pointers within a shared data structure. Modification of non-pointer fields within an RCU-protected data structure is beyond the scope of RCU's correctness guarantees.
	\item Writers wishing to update a pointer in a shared data structure must use the \texttt{rcu\_assign\_pointer} API function. \texttt{rcu\_assign\_pointer} assigns a new value to a pointer in a shared data structure.
	\item Readers must access any data structure pointer field operated on by an \texttt{rcu\_\linebreak[0]assign\_pointer} using \texttt{rcu\_\linebreak[0]dereference}. For example, if the \texttt{next} pointer field in \emph{any} list element of an RCU-protected list has been assigned to by \texttt{rcu\_assign\_\linebreak[0]pointer}, then \emph{all} \texttt{next} pointer fields must be accessed by readers of that list by using \texttt{rcu\_dereference}.
\end{enumerate}

@@

\begin{figure}[ht!] %
\subfloat[Despite correctly obtaining a pointer to a shared data structure using \texttt{rcu\_dereference} within a read-side critical section, properties (2) and (3) tell us that the data pointed to by \texttt{elm} is not guaranteed to exist after the read-side critical section ends.]{\fbox{\usebox\rcuexistencebug}} %

\subfloat[Despite the call to \texttt{rcu\_dereference}, properties (1) and (3) tell us that \texttt{list\_head} is not guaranteed to be equivalent to \texttt{elm} because a writer might have concurrently changed the value of the \texttt{list\_head} pointer. That is, if \texttt{elm} is non-\texttt{NULL}, then the data pointed to by \texttt{elm} is guaranteed to exist for the remainder of the read-side critical section. However, the value of \texttt{list\_head} is \emph{not} guaranteed to remain consistent--even within read-side critical sections--and so \texttt{list\_head} might point to a \texttt{NULL} pointer when \texttt{elm} does not.]{\fbox{\usebox\rcuderefbug}} %

\subfloat[This example shows the code for a writer thread adding a new element into the second position of an RCU-protected list. The code for reader threads is also shown, where readers iterate over the list by traversing \texttt{next} pointers and sum up the \texttt{data} fields of each visited list element. The access of \texttt{elm->next} violates of properties (3) and (7). That is, the next element of the list is only guaranteed to exist within the read-side critical section if the reader executes \texttt{rcu\_dereference(elm->next)}. Directly reading \texttt{elm->next} might yield a \emph{temporarily} valid pointer to the next list element, or it might return an inconsistent value for the pointer that is a mix of the two versions of the pointer.]{\fbox{\usebox\rcuassignderefbug}} %
\caption{\label{fig:rcu_bugs}Example RCU bugs related to misuses of pointers to shared data.}
\end{figure}

@@

@@

>>>>

Read-Copy-Update (RCU) is a scalable, lock-free synchronization mechanism used in the Linux kernel \cite{RCU,RCUInLinux}. When correctly used, RCU permits concurrent reads and writes to shared data structures, but requires that writers synchronize among themselves. However, like other synchronization mechanisms, RCU has a deceptively simple API. Misuses of the RCU API often result in hard-to-find bugs that can be implementation-dependent (i.e. bugs that manifest in one version of RCU but not another) or architecture-specific (e.g. bugs that only exhibit on architectures with a relaxed memory model). For example, RCU bugs sometimes manifest as use-after-free bugs, which can result in silent data corruption (if the underlying memory remains used) or kernel panics.

Unfortunately, approaching RCU from the perspective of finding use-after-free errors is insufficient. A better characterization of this class of RCU bugs is proof-of-existence bugs: a data structure cannot be proven to exist at some potentially allocated memory. Existence in this case relates to semantic reachability. For example, if a list element $E_n$ from an RCU-protected list is reachable from the head of the list by traversing zero-or-more intermediate elements $E_0, ..., E_{n-1}$, then $E_n$ is said to exist. While related, memory leak detection (e.g. via conservate mark and sweep) is also insufficient for ladidadidaaaa TODO TODO TODO

% (assuming that pointer reachability is sufficient to determine semantic reachability in this case).

Granary's \texttt{RCUdbg} tool is designed to catch RCU bugs related to kernel code using pointers to RCU-protected data structures that haven't been proven to exist.

%the associated memory may or may not be free, but the data structure occupying that memory can't be proven to exist. This

%  However, this class of RCU bugs is not well-suited to being detected by a use-after-free memory checker tool
%RCU is notoriously hard to use, despite its relatively simple API. RCU usage bugs, like bugs with other synchronization mechanisms, often stay hidden and only exhibit themselves for very specific thread interleavings/schedules.

% go silently unnoticed because the memory

%While bugs related to mutual exlusion locks might result in easily observable deadlocks, bugs related to RCU typically result in use-after-free errors, which sometimes go silently unnoticed, sometimes result in kernel panics, and most of the time work as expected because the associated data is .

but only if the following properties are satisfied: \begin{enumerate}
	\item Readers must tolerate an inconsistent view of the shared data structure. That is, a reader operating concurrently with a writer might observe either the version of the data before or after the writer's changes have been published.
	\item Readers must operate within \emph{read-side} critical sections, which are delimited by \texttt{rcu\_read\_lock} and \texttt{rcu\_read\_unlock}, respectively.
	\item Within read-side critical sections, readers can only read a shared data structure after obtaining a pointer to that shared data structure using the \texttt{rcu\_dereference} API function. \texttt{rcu\_dereference} operates on a pointer to a shared data structure, and returns a pointer to that can be used to safely access the shared data structure within the remainder of the reader's read-side critical section. Notably, accessing the original data structure and not the dereferenced data structure pointer--even after an \texttt{rcu\_dereference}--is undefined behavior. That is, within a read-side critical section, RCU only guarantees the existence of data structures pointed to by RCU-dereferenced pointers.
	\item Writers must synchronize among themselves, e.g. via mutual exclusion locks or spin locks, before any writes to a shared data structure are performed. We say writes to the shared data structure occur within \emph{write-side} critical sections.
	\item RCU only defines how writers can update pointers within a shared data structure. Modification of non-pointer fields within an RCU-protected data structure is beyond the scope of RCU's correctness guarantees.
	\item Writers wishing to update a pointer in a shared data structure must use the \texttt{rcu\_assign\_pointer} API function. \texttt{rcu\_assign\_pointer} assigns a new value to a pointer in a shared data structure.
	\item Readers must access any data structure pointer field operated on by an \texttt{rcu\_\linebreak[0]assign\_pointer} using \texttt{rcu\_\linebreak[0]dereference}. For example, if the \texttt{next} pointer field in \emph{any} list element of an RCU-protected list has been assigned to by \texttt{rcu\_assign\_\linebreak[0]pointer}, then \emph{all} \texttt{next} pointer fields must be accessed by readers of that list by using \texttt{rcu\_dereference}.
\end{enumerate}

@@

\begin{figure}[ht!] %
\subfloat[\label{fig:rcu_existence_bug}Despite correctly obtaining a pointer to a shared data structure using \texttt{rcu\_dereference} within a read-side critical section, properties (2) and (3) tell us that the data pointed to by \texttt{elm} is not guaranteed to exist after the read-side critical section ends.]{\fbox{\usebox\rcuexistencebug}} %

\subfloat[\label{fig:rcu_deref_bug}Despite the call to \texttt{rcu\_dereference}, properties (1) and (3) tell us that \texttt{list\_head} is not guaranteed to be equivalent to \texttt{elm} because a writer might have concurrently changed the value of the \texttt{list\_head} pointer. That is, if \texttt{elm} is non-\texttt{NULL}, then the data pointed to by \texttt{elm} is guaranteed to exist for the remainder of the read-side critical section. However, the value of \texttt{list\_head} is \emph{not} guaranteed to remain consistent--even within read-side critical sections--and so \texttt{list\_head} might point to a \texttt{NULL} pointer when \texttt{elm} does not.]{\fbox{\usebox\rcuderefbug}} %

\subfloat[\label{fig:rcu_assign_bug}This example shows the code for a writer thread adding a new element into the second position of an RCU-protected list. The code for reader threads is also shown, where readers iterate over the list by traversing \texttt{next} pointers and sum up the \texttt{data} fields of each visited list element. The access of \texttt{elm->next} violates of properties (3) and (7). That is, the next element of the list is only guaranteed to exist within the read-side critical section if the reader executes \texttt{rcu\_dereference(elm->next)}. Directly reading \texttt{elm->next} might yield a \emph{temporarily} valid pointer to the next list element, or it might return an inconsistent value for the pointer that is a mix of the two versions of the pointer.]{\fbox{\usebox\rcuassignderefbug}} %
\caption{\label{fig:rcu_bugs}Example RCU bugs related to misuses of pointers to shared data.}
\end{figure}

The \texttt{RCUdbg} tool detects the the following bugs: \begin{enumerate}
	\item Reads through an RCU-dereferenced pointer outside of a read-side critical section (\Cref{fig:rcu_existence_bug}).
	\item Reads through an RCU-assigned pointer that don't first use \texttt{rcu\_dereference} (\Cref{fig:rcu_deref_bug}).
	\item Reads through fields of RCU-protected data structures that should be accessed via \texttt{rcu\_dereference} (\Cref{fig:rcu_assign_bug}).
	\item Reads through an RCU-dereferenced pointer in the wrong read-side critical section.
	\item Writes through an RCU-dereferenced pointer.
	\item Writes through an RCU-assigned pointer.
\end{enumerate}

\texttt{RCUdbg} uses Granary's address watchpoints feature \cite{BehaveOrBeWatched} to ``taint" addresses participating in RCU-protected data structures.

@@

TODO: JIFL

@@

TODO: SystemTap
